---
title: "The Linear Model (LM)"
author: "T. Florian Jaeger"
date: \today
geometry: margin=2cm
header-includes:
  - \usepackage{animate}
  - \usepackage{booktabs}
  - \usepackage{tabto}
  - \usepackage{soul}
  - \usepackage{xcolor}
  - \usepackage{placeins}
  - \usepackage{lscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \makeatletter\renewcommand{\fps@table}{!ht}\makeatother
  - \setstcolor{red}
  - \usepackage{sectsty}
  - \sectionfont{\color{blue}} 
  - \subsectionfont{\color{blue}}
  - \subsubsectionfont{\color{darkgray}}
  - \usepackage{caption}
  - \usepackage{subcaption}
  - \usepackage{tikz}
  - \usepackage{url}
  - \usetikzlibrary{bayesnet}
output:
  tufte::tufte_handout: default
  fontsize: 10pt
urlcolor: blue
editor_options: 
  markdown: 
    wrap: 80
---

```{r set-options, include=F}
source("constants.R")
```

# Readings and assignments in *preparation* of this class

<!-- In preparation for class, please **read *Regression and other stories* (ROS), -->
<!-- Chapters 6 & 7.1-2**. This text is quite intuitive and assumes comparatively little -->
<!-- background. But if you find yourself getting lost, please stop and first read or -->
<!-- watch one of the many more informal introductions to linear models (e.g., [this -->
<!-- one](https://medium.com/analytics-vidhya/linear-regression-an-intuitive-approach-ded05ed39a87)). -->
<!-- **You should plan for 2h of reading and summarizing your questions in -->
<!-- preparation for this week of the class.** -->

**Optionally:** Before the next meeting, you might find it helpful to read and *work through* this document.
Please note that this document is providing R code only. Some of the steps
described here might have less direct solutions in Matlab or Python, so it is
recommended that you start early.

# About this document

This PDF is generated from an R markdown document (using the \texttt{tufte}
package for styling). If you're interested in understanding the code that
generated the examples presented below, you can [clone the
repo](https://github.com/tfjaeger/Tutorial-GLMM/). Both the PDF and the R
markdown that the PDF is generated from are in the scripts/ folder. You can go
through the document line by line, or make changes and 'knit' a new PDF with the
press of a button.

# The linear model (LM)

Recall the linear model:

\begin{figure}
  \centering
  \tikz{ %
    \node[obs] (outcome) {$y_i$} ; %
    \factor[above=of outcome] {distribution} {left: \textcolor{gray}{$\mathcal{N}$}} {} {}; %
    \node[latent, right=of distribution] (sigma) {$\sigma$} ; %
    \node[det, above=of distribution] (mu) {$\mu_i$} ; %
    \node[obs, above=of mu] (X) {$x_i$} ; %
    \node[latent, right=of X] (beta) {$\beta$} ; %
    % plates
    \plate[inner sep=0.12cm, xshift=-0.06cm, yshift=0.06cm] {plate1} {(mu) (distribution) (X) (outcome)} {$\forall i=1 \ldots N $}; %
    \edge {distribution} {outcome} ; %
    \edge {mu} {distribution} ; %
    \edge {sigma} {distribution} ; %
    \edge {X, beta} {mu} ; %
  }
  \caption{A graphical model visualization of the linear model. Shaded circles are observable variables. White circles are latent variables. These are the variables that constitute the degrees of freedom of the model, the paramters that are fit to the data. Diamonds shapes indicate variables that are determined by other variables. The index $i = 1, ..., N$ refers identifies the observations (usually, the rows) in your data. Note that $\sigma$, unlike $\mu_i$ is assuemd to be constant across observations. This is the assumption of the homogeneity of variance.}\label{fig:lm}

\end{figure}

and the specific LM we fit to a single participant from [Tan & Jaeger (2025)](https://www.sciencedirect.com/science/article/pii/S0010027725001350): 

```{r}
d <- 
  read.csv("../data/Tan-Jaeger.csv") %>%
  # Remove catch trials
  filter(!is.na(Response.Voicing)) %>%
  # Some variable typing
  mutate(
    across(c(ParticipantID, Block, Condition.Exposure), factor),
    Response.t = ifelse(Response.Voicing == "voiceless", 1, 0)) %>%
  rename(VOT = Item.VOT, Condition = Condition.Exposure) %>%
  # making response variable a little easier to understand for non-language scientists
  mutate(Response.isT = ifelse(Response.Voicing == "voiceless", T, F)) 

dd <- 
  d %>% 
  filter(
    Condition == "Shift40",
    Phase == "test") %>%
  filter(ParticipantID == first(ParticipantID)) %>%
  # select only the columns of the data that are relevant for this tutorial
  select(ParticipantID, Condition, Phase, Block, Trial, VOT, Response.isT, Response.RT) 

m <- lm(formula = Response.RT ~ 1 + VOT, data = dd)
summary(m)
```

In this document, we'll go through an example of how this model was fit. We'll start with the perhaps most intuitive fitting algorithm, the *least-square fitting* algorithm. We'll then show how this is equivalent to a more general approach, *likelihood maximization*, which can be used to fit many other types of models.

# Least-square fitting

How did the model fitting algorithm arrive at the estimates and what do they
present? What do the standard errors (SEs) of the estimates represent?
Intuitively, we know that the estimates are the "best-fitting" estimates for the
coefficients. What this means depends on how the goodness of fit is calculated.

One way to find the best-fitting estimates for the coefficients of an LM is to
minimize the total prediction error. For each observation in our data, the
prediction error or *residual* of the model $m$ is defined as the difference
between the predicted value of $y$, ${\rm E_m}[y]$, and the actually observed
value of $y$, i.e., $residual = y - {\rm E_m}[y]$. Since we don't want negative
and positive prediction errors (resulting from over- and under-predicting $y$,
respectively) to cancel each other out, we cannot simply sum those prediction
errors to get an estimate of how well the model fits our data. Instead, we sum
the squared residuals, giving us the residual sum of squares (RSS).[^12] Finding
the best-fitting estimates then becomes the identical to finding the estimates
that minimize the RSS.

[^12]: One could instead use any other function that would result in only
    positive prediction errors (e.g., taking the *absolute* prediction error).
    However, using the squared residuals has both appealing properties, and it
    turns out to be identical---for LMs---to maximizing the likelihood of the
    data.

To better understand what happens when we find the best-fitting estimates by
minimizing the RSS, let's define a few function that help use deliver
predictions just like the final LM presented above does, but for arbitrary
intercept values $a$ and arbitrary slope values $b$ for the effect of VOT We
also define a function that gets the prediction error for each observation in
our data, and a function that sums up the squared prediction error:

```{r, echo=TRUE}
get_prediction <- function(a, b, data = dd) {
  a + b * dd$VOT
}

get_prediction_error <- function(a, b, data = dd) {
  dd$Response.RT - get_prediction(a, b, data = data)
}

get_RSS <- function(a, b, data = dd) {
  sum(get_prediction_error(a, b, data = data)^2)
}
```

Now we can use these functions to visualize the predictions and calculate the
RSSs for different hypotheses about $a$ and $b$. You can use Acrobat Viewer to
animate through the following examples (PDF system preview typically won't
work):

```{r, , fig.show='animate', fig.cap="Raw RTs for one participants across the six test blocks (points), along with prediction line for different intercepts {\\em a} and slopes{\\em b} for the model \\texttt{RT} $\\sim$ \\texttt{1 + VOT}. Compare how the RSS changes, depending on {\\em a} and {\\em b}.", out.width="80%"}
plot_prediction <- function(a, b, data = dd) {
  RSS <- get_RSS(a, b, data)
  
  data %>%
    ggplot(aes(x = VOT, y = Response.RT)) +
    geom_point(alpha = .3) +
    geom_abline(aes(intercept = a, slope = b), color = "blue") +
    annotate(
      geom = "text",
      label = deparse(bquote("E[y]"== .(as.numeric(a))+.(as.numeric(b))*x)),
      parse = T,
      x = 85, y = 5000, hjust = 1, color = "blue", size = 6) +
    annotate(
      geom = "label", 
      label = deparse(bquote(RSS == .(RSS))), 
      parse = T,
      x = 85, y = 0, hjust = 1, vjust = 0, color = "orange", size = 6) +
    xlab("VOT (msecs)") + ylab("Response RT")
}

# plot prediction for best fit
plot_prediction(coef(m)[1], coef(m)[2])

# plot predictions for halved or doubled intercept
plot_prediction(coef(m)[1] / 2, coef(m)[2])
plot_prediction(coef(m)[1] * 2, coef(m)[2])

# plot predictions for zero or inverted slope
plot_prediction(coef(m)[1], coef(m)[2] * 0)
plot_prediction(coef(m)[1], coef(m)[2] * -1)
```

We can also visualize the RSS---i.e., the function we're trying to
minimize---depending on the a wide range of values for $a$ and $b$. This
illustrates two things. First, the estimates provided by the LM at the start of
this section indeed minimize the RSS of the model. Second, RSS are a convex
function of the parameters in the model (here $a$ and $b$), with a single
optimum. This very appealing property of LMs allows for efficient fitting
algorithms that are *guaranteed* to converge within an arbitrary finite number
of steps against the optimal parameters. Basically, it means that going downhill
is bound to get you closer to the optimum, rather than stuck in a local minimum
(as can be the case for more complex optimization problems, e.g,. for many types
of non-linear models).

```{r, fig.cap="RSS for the model \\texttt{RT} $\\sim$ \\texttt{1 + VOT}, depending on its intercept $a$ and slope $b$. The white point shows that location of the estimates given in the LM output at the beginning of this section.", fig.height=4, fig.width=4, out.width="66%"}
# get RSS for a grid of a and b
d.rss <- 
  crossing(
    a = coef(m)[1] * seq(-10, 10, .1),
    b = coef(m)[2] * seq(-10, 10, .1)) %>%
  mutate(RSS = map2_dbl(a, b, ~ get_RSS(.x, .y, dd)))

d.rss %>%
  ggplot(aes(x = a, y = b)) +
  geom_tile(aes(fill = RSS)) +
  geom_point(color = "white", x = coef(m)[1], y = coef(m)[2]) +
  xlab("intercept a") + ylab("slope b of VOT") +
  scale_fill_viridis_b(trans = "log10") +
  coord_cartesian(expand = FALSE) +
  theme(legend.position = "top", legend.key.width=unit(1,"cm"))
```

```{r, fig.height=4, fig.width=4, out.width="66%"}
# Plot RSS at edges of CI
# 
# last_plot() +
#   annotate(
#     geom = "linerange",
#     orientation = "y",
#     y = as.numeric(coef(m)[2]),
#     xmin = qnorm(.025, coef(summary(m))[1,1], coef(summary(m))[1,2]),
#     xmax = qnorm(.975, coef(summary(m))[1,1], coef(summary(m))[1,2]),
#     color = "white") +
#   annotate(
#     geom = "linerange",
#     x = as.numeric(coef(m)[1]),
#     ymin = qnorm(.025, coef(summary(m))[2,1], coef(summary(m))[2,2]),
#     ymax = qnorm(.975, coef(summary(m))[2,1], coef(summary(m))[2,2]),
#     color = "white") +
#   annotate(
#     geom = "text",
#     x = as.numeric(coef(m)[1]),
#     y = qnorm(.025, coef(summary(m))[2,1], coef(summary(m))[2,2]),
#     label = round(d.rss[
#       d.rss$a == d.rss$a[which.min(abs(d.rss$a - qnorm(.025, coef(summary(m))[2,1], coef(summary(m))[2,2])))] &
#       d.rss$b == d.rss$b[which.min(abs(d.rss$b - as.numeric(coef(m)[1])))],]$RSS),
#     vjust = 1, color = "white") +
#   annotate(
#     geom = "text",
#     x = as.numeric(coef(m)[1]),
#     y = qnorm(.975, coef(summary(m))[2,1], coef(summary(m))[2,2]),
#     label = round(d.rss[
#       d.rss$a == d.rss$a[which.min(abs(d.rss$a - qnorm(.975, coef(summary(m))[2,1], coef(summary(m))[2,2])))] &
#       d.rss$b == d.rss$b[which.min(abs(d.rss$b - as.numeric(coef(m)[1])))],]$RSS),
#     vjust = 0, color = "white") +
#   annotate(
#     geom = "text",
#     x = qnorm(.025, coef(summary(m))[1,1], coef(summary(m))[1,2]),
#     y = as.numeric(coef(m)[2]),
#     label = round(d.rss[
#       d.rss$a == d.rss$a[which.min(abs(d.rss$a - qnorm(.025, coef(summary(m))[1,1], coef(summary(m))[1,2])))] &
#       d.rss$b == d.rss$b[which.min(abs(d.rss$b - as.numeric(coef(m)[2])))],]$RSS),
#     hjust = 1, color = "white") +
#   annotate(
#     geom = "text",
#     x = qnorm(.975, coef(summary(m))[1,1], coef(summary(m))[1,2]),
#     y = as.numeric(coef(m)[2]),
#     label = round(d.rss[
#       d.rss$a == d.rss$a[which.min(abs(d.rss$a - qnorm(.975, coef(summary(m))[1,1], coef(summary(m))[1,2])))] &
#       d.rss$b == d.rss$b[which.min(abs(d.rss$b - as.numeric(coef(m)[2])))],]$RSS),
#     hjust = 0, color = "white") +
#   coord_cartesian(
#     xlim = 
#       qnorm(.5, coef(summary(m))[1,1], coef(summary(m))[1,2]) +
#       (range(
#         qnorm(.025, coef(summary(m))[1,1], coef(summary(m))[1,2]),
#         qnorm(.975, coef(summary(m))[1,1], coef(summary(m))[1,2])) - 
#          qnorm(.5, coef(summary(m))[1,1], coef(summary(m))[1,2])) * 2,
#     ylim = 
#       qnorm(.5, coef(summary(m))[2,1], coef(summary(m))[2,2]) + 
#       (range(
#         qnorm(.025, coef(summary(m))[2,1], coef(summary(m))[2,2]),
#         qnorm(.975, coef(summary(m))[2,1], coef(summary(m))[2,2])) - 
#          qnorm(.5, coef(summary(m))[2,1], coef(summary(m))[2,2])) * 2)
```

Here's an example of how one can use a general optimizer to find the
best-fitting estimates for an LM. The example uses R's \texttt{optim()} function
but you can try to implement it in your preferred programming language.[^13] You
can compare its output to the estimates reported at the start of the section:

[^13]: By default, \texttt{optim()} uses Quasi-Newtonian optimization. This
    works just fine for the present purpose.

```{r, echo=T}
# let's store the optimizer's steps in a data.frame
d.trace <- tibble(.rows = 0)

o <- 
  optim(
    # Starting values for a and b
    par = c(0, 0), 
    fn = function(x) {
      rss <- get_RSS(x[1], x[2]);
      # Uncomment this line to see how optim() searches the parameter space
      # message(paste0("a = ", x[1], ", b = ", x[2], "; RSS = ", rss, "\n")) 
      # <<- assigns values to *global* variables (outside the function)
      d.trace <<- d.trace %>% bind_rows(tibble(a = x[1], b = x[2], RSS = rss))
      return(rss)
    })

# print optimal parameters found by optim()
print(o$par)
```

```{r, fig.margin=T, fig.height=4, fig.width=4, out.width="100%", fig.cap="Same as above but zoomed in and showing the path of the optimizer superimposed. Labels show step of optimizer", message=F, warning=F}
# add row numbers
d.trace %<>%
  mutate(step = as.numeric(row.names(.)))

last_plot() + 
  geom_path(
    data = d.trace, 
    color = "lightgray") +
  geom_label(
    data = d.trace %>% slice(c(1, 30, 60, 90, 120)),
    aes(label = step),
    color = "darkgray") +
  coord_cartesian(xlim = range(d.trace$a), ylim = range(d.trace$b))
```

### Want to challenge yourself? Apply what you learned

While minimization of RSS works well for LMs, there is a more generally
applicable approach that yields the same results for LMs but also can be used
for the fitting of many others types of models, *likelihood maximization*. Under
this approach, we find the best-fitting estimates that maximize the likelihood
of the observed outcomes $y$. Recall that the LM assumes $y$ to be normally
distributed around the predicted ${\rm E}[y]$. This means we can calculate the
*likelihood* of each observation under the hypothesized normal distribution.
This normal distribution is a function of the hypothesized values for all the
model's coefficients ($a$ and $b$ in the example above) *and* the value
hypothesized for $\  sigma$. For any hypothesis about $a$, $b$, and $\sigma$, we
can get the density of an observation $y$ under $\mathcal{N}(a + bx, \sigma)$.
By multiplying the densities of all observations, we can get the data likelihood
under the hypothesized values for $a$, $b$, and $\sigma$. In practice, we
instead sum the log-transformed likelihoods. This is numerically more stable,
and the $x$ that maximizes $\log(f(x))$ is the same that maximizes $f(x)$.

-   Define a function that calculates the sum of log-densities for the model
    \texttt{RT} $\sim$ \texttt{1 + VOT} given a set of arguments $a$, $b$, and
    $\sigma$.[^14]

-   Make a figure like the one shown above for RSS. For this figure set $\sigma$
    to the $\sigma$ from the LM at the start of this section.[^15] We don't
    expect the likelihood values to be identical to the RSS values---they are on
    completely different scales. But do the surfaces have similar shapes
    (curvatures)?

-   To further understand the relationship between OLS and ML fitting of linear
    regression with normally distributed residuals, see [this presentation,
    slides 99-122](https://faculty.washington.edu/cadolph/mle/topic2.p.pdf]
    (it's a lot of slides but that's just because each derivation step is
    presented as a separate slide).

[^14]: In R, \texttt{dnorm()} will give you the density of an observation or
    vector of observations $y$ under a Normal with the specified mean and
    standard deviations. Use the optional argument \texttt{log = TRUE} to get
    log densities.

[^15]: In R, use \texttt{sigma()} to extract $\sigma$ from an LM.

# Session info

```{r session_info, include=TRUE, echo=TRUE, results='markup'}
devtools::session_info()
```
